### 2018-07-18 pipeline.tex
Extend chapter 3

Several proteomics pipelines are available on the internet under different licensing conditions. Many are released as closed-source commercial software, where information on how the program works is kept from the user. This is a serious drawback as it hinders the study of the implemented models and its customisation. Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, good broad base or format exchangeability. MaxQuant, a freeware but closed-source and monolithic proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.

@@

\begin{figure}[!h]
\begin{subfigure}{.9\textwidth}
  \centering
    \caption*{A}
\includegraphics[width=.9\linewidth]{peptide_profile}
\end{subfigure}
\bigskip

@@

The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} (\ac{log2FC}) across condition H and condition L. During quantification, it is important to take into account as many of the effects influencing protein quantification. In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished. The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed. Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{curves_plot}
\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available. \textbf{Comp+Rob} is the pipeline presented in the previous sections. \textbf{MQ+LFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine. Finally, MQ+Rob blends MaxQuant upstream analysis with the MSqRob quantification engine. \textbf{A, B} ROC curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively. \textbf{C, D} PR curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively.}
\label{fig:roc_curves}
\end{figure}

@@

\begin{figure}[!h]
\centering
\begin{subfigure}{.40\textwidth}
\caption*{A}
\includegraphics[width=.9\textwidth]{vennDiagram}
\end{subfigure}
\begin{subfigure}{.52\textwidth}
\caption*{B}
\includegraphics[width=.95\textwidth]{correlation_plot}
\end{subfigure}
\caption{Agreement between the different pipelines. \textbf{A} The Venn diagram displays the number of \textit{E. coli} proteins found to be differentially abundant by each pipeline. \textbf{B} The log2FC estimated in all three pipelines for the 121 shared proteins is plotted by pairs to evaluate their correlation.}
\label{fig:venn_cor}
\end{figure}

In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn_cor} A). Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins. The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detectd by MQ+Rob too. Remarkably, 517 proteins were detected by MQ+Rob alone, which further confirmed its resolving power. Analyzing the log2FC pairwise-correlation (see figure \ref{fig:venn_cor} B) reveals that the highest correlation was found between pipelines using MaxQuant as upstream spectra processor, as expected. Even though the observed correlation was low, the estimated log2FC was always greater than 1. The MQ+Rob pipeline systematically estimated log2FV values greater than the other 2.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{combined_plot}
\caption{Overview over the different pipelines performance. Quantification results are segregated by the source organism. \textbf{A} Histogram of the \ac{log2FC} estimates. \textbf{B} Volcano plot, where every protein group is represented by a dot and the coordinates map to its \ac{log2FC} on x and the minus logarithm of the q-value on y. \textbf{C} Boxplots of the minus logarithm of the q-value distribution.}
\label{fig:combined_plot}
\end{figure}

@@

MSqRob treats the fractions as a random effect, as explained in section \ref{subsec:quantification}. As a consequence, the statistical model in MSqRob can be overwhelmed by the resulting inconsistency, and eventually provide null estimates of the ratio. This is clearly manifested in figure \ref{fig:combined_plot} as the bins at \ac{log2FC} of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis. While no \textit{E. coli} proteins fell in this category in the MQ-MSqRob pipeline, many did in the Comp+Rob, proving room for improvement in this aspect.

@@

The application of mass spectrometry to proteomics is a very hot topic in research, driving innovations in the way the different components of the spectrometer work. One recent example is the development of the timsTOF\texttrademark\xspace Pro (Trapped Ion Mobility Spectrometry-Time Of Flight) system by Bruker, which supports PASEF (Parallel Accumulation and SErial Fragmentation) spectrometry \footnote{\href{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}}. Notwithstanding, it reveals the immature state of the technology. Slight inconsistencies in the protein extraction and digestion, bias in the peptide ionization, presence of unexpected PTMs, and detector saturation and insensitivity, remain as problems contributing to the eventual measurement variability characteristic of mass spectrometry \cite{Piehowski2013}. Thus, adequate cleaning of the peptide dataset, and care to ignore outliers is preeminent. Indeed, the input file for the MQ+LFQ consisted of a preprocessed file which removed the most problematic proteins \cite{Cox2014}, which explains why more than 500 extra proteins could be quantified when using MSqRob (see figure \ref{fig:venn_cor} A).

@@

>>>>

Several proteomics pipelines are available on the internet under different licensing conditions. Many are released as closed-source commercial software, where information on how the program works is kept from the user. This is a serious drawback as it hinders the study of the implemented models and its customisation. Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, good broad base or format exchangeability. MaxQuant, a freeware but closed-source, monolithic proteomics analysis suite \cite{Cox2008}, is developed for Windows and has only very recently been ported to Linux \cite{Sinitcyn2018}. Nevertheless has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.

@@

\begin{figure}[H]
\begin{subfigure}{.9\textwidth}
  \centering
    \caption*{A}
\includegraphics[width=.9\linewidth]{peptide_profile}
\end{subfigure}
\bigskip

@@

The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $\log_2$(ratios), or \textit{log2-fold-changes} (\ac{log2FC}) across condition H and condition L. During quantification, it is important to take into account as many of the effects influencing protein quantification. In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished. The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed. Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{curves_plot}
\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available. \textbf{Comp+Rob} is the pipeline presented in the previous sections. \textbf{MQ+LFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine. Finally, MQ+Rob blends MaxQuant upstream analysis with the MSqRob quantification engine. \textbf{A, B} ROC curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively. \textbf{C, D} PR curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively.}
\label{fig:roc_curves}
\end{figure}

@@

\begin{figure}[H]
\centering
\begin{subfigure}{.40\textwidth}
\caption*{A}
\includegraphics[width=.9\textwidth]{vennDiagram}
\end{subfigure}
\begin{subfigure}{.52\textwidth}
\caption*{B}
\includegraphics[width=.95\textwidth]{correlation_plot}
\end{subfigure}
\caption{Agreement between the different pipelines. \textbf{A} The Venn diagram displays the number of \textit{E. coli} proteins found to be differentially abundant by each pipeline. \textbf{B} The log2FC estimated in all three pipelines for the 121 shared proteins is plotted by pairs to evaluate their correlation.}
\label{fig:venn_cor}
\end{figure}

In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn_cor} A). Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins. The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detectd by MQ+Rob too. Remarkably, 517 proteins were detected by MQ+Rob alone, which further confirmed its resolving power. Analyzing the log2FC pairwise-correlation (see figure \ref{fig:venn_cor} B) reveals that the highest correlation was found between pipelines using MaxQuant as upstream spectra processor, as expected. Even though the observed correlation was low, the estimated log2FC was always greater than 1. The MQ+Rob pipeline systematically estimated \ac{log2FC} values greater than the other 2.

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{combined_plot}
\caption{Overview over the different pipelines\textquotesingle performance. Quantification results are segregated by the source organism. \textbf{A} Histogram of the \ac{log2FC} estimates. \textbf{B} Volcano plot, where every protein group is represented by a dot and the coordinates map to its \ac{log2FC} on x and the minus logarithm of the q-value on y. \textbf{C} Boxplots of the minus logarithm of the q-value distribution.}
\label{fig:combined_plot}
\end{figure}

@@

MSqRob treats the fractions as a random effect, as explained in section \ref{subsec:quantification}. As a consequence, the statistical model in MSqRob can be overwhelmed by the inconsistency resulting from a 24-plex sample fractionation, leading to null \ac{log2FC} estimates for many \textit2{E. coli} proteins. This is clearly manifested in figure \ref{fig:combined_plot} as the bins at \ac{log2FC} of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis. While no \textit{E. coli} proteins fell in this category in the MQ-MSqRob pipeline, many did in the Comp+Rob, proving room for improvement in this aspect.

@@

The application of mass spectrometry to proteomics is a very hot topic in research, driving innovations in the way the different components of the spectrometer work. One recent example is the development of the timsTOF\texttrademark\xspace Pro (Trapped Ion Mobility Spectrometry-Time Of Flight) system by Bruker, which supports PASEF (Parallel Accumulation and SErial Fragmentation) spectrometry \footnote{\href{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}}. This technology adds  an extra dimension peptide separation process, leading to cleaner spectra.
Notwithstanding, the developments also reveal the immature state of the technology. Slight inconsistencies in the protein extraction and digestion, bias in the peptide ionization, presence of unexpected PTMs, and detector saturation and insensitivity, remain as problems contributing to the eventual measurement variability characteristic of mass spectrometry \cite{Piehowski2013}. Thus, adequate cleaning of the peptide dataset, and care to ignore outliers is preeminent. Indeed, the input file for the MQ+LFQ consisted of a preprocessed file which removed the most problematic proteins \cite{Cox2014}, which explains why more than 500 extra proteins could be quantified when using MSqRob (see figure \ref{fig:venn_cor} A).

@@

\end{document}
